Covering Your Basis: Model Agnostic Probes of the Binary Black Hole Population

### 2022-09-26 src/tex/intro.tex
remove comment"
add more conc

Observations of gravitational waves (GWs) from compact binary coalescences (CBCs) is becoming a regular occurrence, 
recently producing catalogs of 90 such detection \cite{GWTC1,gwtc2,GWTC3}. With this growing catalog size, we are steadily approaching a more complete  
understanding of the population properties CBCs, just 7 years after the first observation \cite{o1o2_pop,o3a_pop,o3b_astro_dist}. 
In the coming fourth observing run of the LIGO-Virgo-KAGRA collaboration detectors \citep{aLIGO,aVIRGO}, with the increased sensitivities, 
there could be a detection as often as every day \cite{LVK_prospects}. This large influx of incoming data will drastically increase what we can learn by combining all observations
in order to studying the compact object population. The history of these dense objects is imprinted on the properties of the population 
of CBCs, allowing us to learn about their formation and evolution by the last few seconds of gravitation radiation emitted. The population distibutions 
of merging compact objects inferred from gravitational-wave observations has been shown to be parituclar useful in order to, measure cosmological parameters \citep{Farr_2019HUB,gwtc3_cosmo}, 
constrain modified gravitational wave propagation \citep{ModGWProp,ModGWProp2}, constrain a running Planck mass \citep{Lagos_runningPlanckMass}, look for evidence of ultralight bosons through superradiance \citep{Ng_Boson2021,GWTC2_superradiance_Ng}, 
constrain stellar nuclear reaction rates \citep{Farmer_2019,Farmer_2020}, look for primordial black holes \citep{Ng_2021,KenNgPBH2022}, and constrain the neutron star equation of state \citep{Golomb_EOS}. 
Through a better understanding of the mass, spin, and redshift distributions of compact binaries that will come with the increased catalog size, 
one can probe a wide range of different physical phenoma with even greater fidelity. \bruce{Cite others??}

Studies using flexible methods to analyze the binary black hole (BBH) mass distribution with GWTC-2 \citep{Tiwari_2021_b,Edelman_2022ApJ} found signs of structure 
beyond the reported $m_1\sim35\,\msun$ feature by \citet{Talbot_2019, o3a_pop}. The has evidence for this structure, such as the peak at $m_1\sim10\,\msun$, has only increased
with GWTC-3 \cite{o3b_astro_dist,Tiwari_2022ApJ}. This substructure was first confidently found with mutliple different flexible or non-parametric models that aim 
to be data-oriented with little bias on the functional form. Conversely the other canonical approach is to use simple parametric models that aim to describe the data
in the simplest way, building in astrophyscially motivated priors, followed by sequentially adding complexities only as the data demands it. While in the current environment 
this simple approach is clearly more powerful, as we go forward into the data-rich catalog era flexible models can be a useful complement.
Non-parametric or semi-parametric methods are particularly useful at looking for unexpected features in the data, or finding where 
our simpler parametric models may be missing features in the data. Development and robust teting of these types of methods in the short term will provide
knock dowm effects later down the line as the size of our catalogs increase and we enter a data-dominated regime rather than prior/model dominated.

>>>>

Observations of gravitational waves (GWs) from compact binary coalescences (CBCs) is becoming a regular occurrence, 
recently producing catalogs of 90 such detection \cite{GWTC1,gwtc2,GWTC3}. With this growing catalog size, we are steadily approaching a more complete  
understanding of the population properties CBCs, just 7 years after the first observation \cite{o1o2_pop,o3a_pop,o3b_astro_dist}. 
In the coming fourth observing run of the LIGO-Virgo-KAGRA collaboration detectors \citep{aLIGO,aVIRGO}, with the increased sensitivities, 
there could be a detection as often as every day \cite{LVK_prospects}. This large influx of incoming data will drastically increase what we can learn by combining all observations
in order to studying the compact object population. The history of these dense objects is imprinted on the properties of the population 
of CBCs, allowing us to learn about their formation and evolution by the last few seconds of gravitation radiation emitted. The population distibutions 
of merging compact objects inferred from gravitational-wave observations has been shown to be parituclar useful in order to, measure cosmological parameters \citep{Farr_2019HUB,gwtc3_cosmo}, 
constrain modified gravitational wave propagation \citep{ModGWProp,ModGWProp2}, constrain a running Planck mass \citep{Lagos_runningPlanckMass}, look for evidence of ultralight bosons through superradiance \citep{Ng_Boson2021,GWTC2_superradiance_Ng}, 
constrain stellar nuclear reaction rates \citep{Farmer_2019,Farmer_2020}, look for primordial black holes \citep{Ng_2021,KenNgPBH2022}, and constrain the neutron star equation of state \citep{Golomb_EOS}. 
Through a better understanding of the mass, spin, and redshift distributions of compact binaries that will come with the increased catalog size, 
one can probe a wide range of different physical phenoma with even greater fidelity.

Studies using flexible methods to analyze the binary black hole (BBH) mass distribution with GWTC-2 \citep{Tiwari_2021_b,Edelman_2022ApJ} found signs of structure 
beyond the reported $m_1\sim35\,\msun$ feature by \citet{Talbot_2018, o3a_pop}. The has evidence for this structure, such as the peak at $m_1\sim10\,\msun$, has only increased
with GWTC-3 \cite{o3b_astro_dist,Tiwari_2022ApJ}. This substructure was first confidently found with mutliple different flexible or non-parametric models that aim 
to be data-oriented with little bias on the functional form. Conversely the other canonical approach is to use simple parametric models that aim to describe the data
in the simplest way, building in astrophyscially motivated priors, followed by sequentially adding complexities only as the data demands it. While in the current environment 
this simple approach is clearly more powerful, as we go forward into the data-rich catalog era flexible models can be a useful complement.
Non-parametric or semi-parametric methods are particularly useful at looking for unexpected features in the data, or finding where 
our simpler parametric models may be missing features in the data. Development and robust teting of these types of methods in the short term will provide
knock dowm effects later down the line as the size of our catalogs increase and we enter a data-dominated regime rather than prior/model dominated.

### 2022-09-26 src/tex/results.tex
add more conc

We use hierarchical Bayesian inference (see Appendix \ref{sec:hierarchical_inference}) to simultaneosly infer the astrophysical mass, spin and redshift distributions of 
a binary black holes (BBHs) given a catalog of gravitational wave observations. We parameterize the binaries' masses with the primary (defined as the more massive component) mass 
and the mass ratio -- defined as $q=\frac{m_2}{m_1}$, bounded in the range from 0 to 1. We model 4 of the 6 total 
spin degrees of freedom of a binary merger with each component spin magnitude $a_1$ and $a_2$, along with the cosine of the tilt angle of each component, 
$\cos{\theta_1}$ and $\cos{\theta_2}$. The tilt angle is defined as the angle between each components' spin vector and the binaries orbital angular mommentum vector. 
We additionally will fit a population model on the redshift or luminosity distance distribution of BBHs. We assume a $\Lambda\mathrm{CDM}$ cosmology using the 
parameters from the Planck 2015 results \cite{Planck2015}, which defines an analytical mapping between each event's inferred
luminosity distance and it's redshift. While we choose to use MSpline distibutions to model the primary mass and spin distributions, we 
model the mass ratio distribution with a power law, i.e. $p(q | m_1, m_\mathrm{min}, \beta) \propto q^{\beta} \Theta(qm_1 - m_\mathrm{min}) \Theta(m_1 - qm_1)$. 
Where $\beta$ is the power law index and $\Theta$ is the Heaviside step function which ensures $m_2$ is within the allowed range, [$m_\mathrm{min}$, $m_1$] \cite{Talbot_2019,o1o2_pop,o3a_pop}.
We model the evolution of the merger rate with redshift with the \textsc{PowerlawRedshift} model \cite{Fishbach_2018redshift}, which is desribed by 
$p(z|\lambda_z)\propto \frac{dV_c}{dz}(1+z)^{\lambda_z-1}$. We describe our specific prior choices on the hyper-parameters and each model 
in more detail in Appendix \ref{sec:modelpriors}. We use the same cut on the recent GWTC-3 catalog as was done in the LVK's accompanying population analysis, 
keeping all 70 possibly BBH mergers with false alarm rates less than 1 per year \cite{GWTC3,o3b_astro_dist}. We choose to emit the currently not well understood event, 
GW190814 \cite{190814disc}, as it was concluded to be an outlier of the rest of the BBH population in both GWTC-2 and GWTC-3 \cite{o3a_pop,o3b_astro_dist,Essick_2022}. 
This leaves us with a catalog of 69 BBH mergers observerd oveer a period of about 2 years, from which we want to infer its population properties. 
We start by looking towards the mass distribution.

@@

We apply our newly constructed models to the BBH analysis that was done in the astrophysical population companion paper to GWTC-3. 
Figure \ref{fig:mass_distribution} shows how the MSpline model (red) found consistent features in the mass disribution with 
what was inferred with the \textsc{PowerlawPeak} and \textsc{PowerlawSpline} mass models \cite{Talbot_2019,o3a_pop,Edelman_2022ApJ,o3b_astro_dist}. 
In particular our MSpline model finds a peak in merger rate density at both 10 and 35 soloar masses, agreeing with those reported using the same dataset in \citet{o3b_astro_dist}. 
We also find a significantly less declining mass distribution that has considerably larger support for higher mass mergers when 
compared to the other power-law based models, finding that 75\% percent of primary masses below \result{$\CIPlusMinus{\macros[MassDistribution][MSpline][m_75percentile]}$}, 
at 90\% credibility, compared to \result{$\CIPlusMinus{\macros[MassDistribution][PLSpline][m_75percentile]}$}, found with the similar \textsc{PowerlawSpline} model. 
An interesting consequence of this is that we do not find a confidently evolving merger rate with redshift, as inferred with power-law based mass models in \citet{o3b_astro_dist}. 
There is a strong correlation between the primary mass power law index, $\alpha$, and the redshift power-law index, $\lambda_z$. When one infers a steeper mass distribution -- 
the inferred rate must drastically increase with resdshift to explain the highest mass events in the catalog. However, when one infers a 
flatter distribution like, as the MSPline did in figure \ref{fig:mass_distribution}, the higher mass mergers are consistent with 
the local merger rate denstiy as a function of mass given the selection effects. With our more flexible mass model, we find that the merger rate 
is consistent with being unifom with co-moving volume, with $\lambda_z = $\result{$\CIPlusMinus{\macros[MSplineIIDCompSpins][lamb]}$}, compared with 
\citet{o3b_astro_dist}'s reported value -- $\lambda_z = $\result{$\CIPlusMinus{\macros[PLPeak][lamb]}$} with the \textsc{PowerlawPeak} mass model.

>>>>

We use hierarchical Bayesian inference (see Appendix \ref{sec:hierarchical_inference}) to simultaneosly infer the astrophysical mass, spin and redshift distributions of 
a binary black holes (BBHs) given a catalog of gravitational wave observations. We parameterize the binaries' masses with the primary (defined as the more massive component) mass 
and the mass ratio -- defined as $q=\frac{m_2}{m_1}$, bounded in the range from 0 to 1. We model 4 of the 6 total 
spin degrees of freedom of a binary merger with each component spin magnitude $a_1$ and $a_2$, along with the cosine of the tilt angle of each component, 
$\cos{\theta_1}$ and $\cos{\theta_2}$. The tilt angle is defined as the angle between each components' spin vector and the binaries orbital angular mommentum vector. 
We additionally will fit a population model on the redshift or luminosity distance distribution of BBHs. We assume a $\Lambda\mathrm{CDM}$ cosmology using the 
parameters from the Planck 2015 results \cite{Planck2015}, which defines an analytical mapping between each event's inferred
luminosity distance and it's redshift. While we choose to use MSpline distibutions to model the primary mass and spin distributions, we 
model the mass ratio distribution with a power law, i.e. $p(q | m_1, m_\mathrm{min}, \beta) \propto q^{\beta} \Theta(qm_1 - m_\mathrm{min}) \Theta(m_1 - qm_1)$. 
Where $\beta$ is the power law index and $\Theta$ is the Heaviside step function which ensures $m_2$ is within the allowed range, [$m_\mathrm{min}$, $m_1$] \cite{Talbot_2018,o1o2_pop,o3a_pop}.
We model the evolution of the merger rate with redshift with the \textsc{PowerlawRedshift} model \cite{Fishbach_2018redshift}, which is desribed by 
$p(z|\lambda_z)\propto \frac{dV_c}{dz}(1+z)^{\lambda_z-1}$. We describe our specific prior choices on the hyper-parameters and each model 
in more detail in Appendix \ref{sec:modelpriors}. We use the same cut on the recent GWTC-3 catalog as was done in the LVK's accompanying population analysis, 
keeping all 70 possibly BBH mergers with false alarm rates less than 1 per year \cite{GWTC3,o3b_astro_dist}. We choose to emit the currently not well understood event, 
GW190814 \cite{190814disc}, as it was concluded to be an outlier of the rest of the BBH population in both GWTC-2 and GWTC-3 \cite{o3a_pop,o3b_astro_dist,Essick_2022}. 
This leaves us with a catalog of 69 BBH mergers observerd oveer a period of about 2 years, from which we want to infer its population properties. 
We start by looking towards the mass distribution.

@@

We apply our newly constructed models to the BBH analysis that was done in the astrophysical population companion paper to GWTC-3. 
Figure \ref{fig:mass_distribution} shows how the MSpline model (red) found consistent features in the mass disribution with 
what was inferred with the \textsc{PowerlawPeak} and \textsc{PowerlawSpline} mass models \cite{Talbot_2018,o3a_pop,Edelman_2022ApJ,o3b_astro_dist}. 
In particular our MSpline model finds a peak in merger rate density at both 10 and 35 soloar masses, agreeing with those reported using the same dataset in \citet{o3b_astro_dist}. 
We also find a significantly less declining mass distribution that has considerably larger support for higher mass mergers when 
compared to the other power-law based models, finding that 75\% percent of primary masses below \result{$\CIPlusMinus{\macros[MassDistribution][MSpline][m_75percentile]}$}, 
at 90\% credibility, compared to \result{$\CIPlusMinus{\macros[MassDistribution][PLSpline][m_75percentile]}$}, found with the similar \textsc{PowerlawSpline} model. 
An interesting consequence of this is that we do not find a confidently evolving merger rate with redshift, as inferred with power-law based mass models in \citet{o3b_astro_dist}. 
There is a strong correlation between the primary mass power law index, $\alpha$, and the redshift power-law index, $\lambda_z$. When one infers a steeper mass distribution -- 
the inferred rate must drastically increase with resdshift to explain the highest mass events in the catalog. However, when one infers a 
flatter distribution like, as the MSPline did in figure \ref{fig:mass_distribution}, the higher mass mergers are consistent with 
the local merger rate denstiy as a function of mass given the selection effects. With our more flexible mass model, we find that the merger rate 
is consistent with being unifom with co-moving volume, with $\lambda_z = $\result{$\CIPlusMinus{\macros[MSplineIIDCompSpins][lamb]}$}, compared with 
\citet{o3b_astro_dist}'s reported value -- $\lambda_z = $\result{$\CIPlusMinus{\macros[PLPeak][lamb]}$} with the \textsc{PowerlawPeak} mass model.
